Dilution Experiment, negative ion mode, intracellular metabolites, WGNCA analysis
This analysis is done after the FT-LC data and TSQ data have been processed
May 30, 2017
Krista Longnecker
Woods Hole Oceanographic Institution

========================================================

Code originally based on the WGCNA tutorials available here:
https://labs.genetics.ucla.edu/horvath/htdocs/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/

```{r setup, eval=TRUE}

library(WGCNA)
library(vegan)
library(ggplot2)
library(ggdendro)
library(flashClust) 
library(xlsx) #need this to read in the xlsx file used for the metadata

options(stringsAsFactors = FALSE)

disableWGCNAThreads() #this is needed to use this code in RStudio

```

```{r FileNames, eval = TRUE}
#setup file names ahead of time, will be used to save and export data

NameOfFile = 'DilExp_WGCNA_negOmni_log2.Rimage'
WriteCSVname = 'DilExp_WGCNA_negOmni_log2.csv'
writeColorsName = 'DilExp_WGCNA_tableColors_log2.csv'

```

Read in the datasets. FT-LC data are untargeted metabolites. TSQ data are the concentrations from select known metabolites (targeted metabolomics)


```{r readInData, eval=TRUE}

#FT-LC data
wDir = "pathToData/FTLCdata/"
fData = read.csv(paste0(wDir,"DilutionExperiment_FTLCdata.csv"))

datExp0 = as.data.frame(t(fData[,-c(1:3)])) #note the transpose here

names(datExp0) = fData$feature
rownames(datExp0) = names(fData[,-c(1:3)])

#prune the data a little - mzRT features must be in multiple samples
gsg = goodSamplesGenes(datExp0,verbose = 3);
datExp0 = datExp0[gsg$goodSamples,gsg$goodGenes]

#TSQ data
wDir <- "pathToData/TSQdata/"
traitData = read.csv(paste0(wDir,"DilutionExperiment_TSQdata.csv"))

dim(traitData)
names(traitData)

#this will match the rows (so I don't need to sort the input)
dSamples = rownames(datExp0)
traitRows = match(dSamples,traitData$Observations) 


#datTraits = traitData[traitRows,-1] #skip the first column which is the sample names
datTraits_all = traitData[traitRows,-c(1)] #skip names (added by R)

#prune out the columns with all zeros...(thanks to Stack Overflow)
tpa = decostand(datTraits_all,method = 'pa')
datTraits_all = datTraits_all[,colSums(tpa)>3]
rm(tpa)

#normalize the trait data so they also all scale 0 to 1
datTraits = decostand(datTraits_all,method = "range")

```


```{r readAnnotations, eval = TRUE}
#label information for each of the mzRT features
#names(datExp0) #this gives all probeIDs, in our case this is mz_rt
#names(datExp0)[moduleColors=="lightyellow"] #this returns probeIDs belonging to one color

annot = read.csv("DilutionExperiment_FTLCdata_annotation.csv") #
dim(annot)
names(annot)

probes = names(datExp0)
probes2annot= match(probes,annot$feature)

```

Have some options for the distance matrix:
```{r calcDistance,eval=TRUE}

#data pre-processing
dataRel = decostand(x=datExp0,method = "log",logbase = 2)
r = vegdist(dataRel,method = "bray") #requires the vegan package

sampleTree = flashClust(r,method = "ward")
sizeGrWindow(5,5)
par(cex=1.5)
par(mar = c(2,2,2,3))
plot(as.dendrogram(sampleTree),horiz=T)

```



```{r combinePlots, eval=TRUE}
traitColors = numbers2colors(datTraits,signed=FALSE)

plotDendroAndColors(sampleTree,traitColors,groupLabels = names(datTraits),
                    main = 'sample dendrogram and heatmap')

```


```{r determinePower,eval=TRUE}

#choose a set of soft-thresholding powers
powers = c(seq(from=1,to=20,by=1))

#call the network topology analysis function
sft = pickSoftThreshold(dataRel,powerVector = powers,verbose = 5,corFnc = cor,
                        networkType = 'signed')

# sizeGrWindow(9,5)
par(mfrow =c(1,2))
cex1 = 0.9;
# 
plot(sft$fitIndices[,1],-sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab = "Soft  threshold (power)", 
     ylab = "scale free topology model fit signed r^2",type = "n", 
     main = paste("scale independence"))

text(sft$fitIndices[,1],-sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels = powers,cex=cex1,col="red")

abline(h=0.9,col = "red")

plot(sft$fitIndices[,1],sft$fitIndices[,5],xlab = "Soft threshold (power)", 
     ylab = "mean connectivity",type = "n", main = paste("mean connectivity"))
text(sft$fitIndices[,1],sft$fitIndices[,5],labels = powers,cex=cex1,col="red")


```

In this next section, I use the merging threshold and soft power found in the previous code chunk. This is specific to our data, and would need to be changed for other data.

```{r setVariables, eval = TRUE}

# remember: threshold of 0.25 corresponds to a correlation of 0.75
mergingThresh = 0.35
softPower = 12

#select some variables of interest for plotting later:
oneTrait = as.data.frame(datTraits$DMSP)
names(oneTrait)="ectoine"

secondTrait = as.data.frame(datTraits$proline)
names(secondTrait) = "DMSP"

```

DoLazyAnalysis essentially uses the default parameters (See below to select each one)

```{r DoLazyAnalysis, eval = TRUE}

#set maxBlockSize to exceed the number of mzRT features in my dataset

net = blockwiseModules(datExp0,corType="pearson",
  maxBlockSize=6000,networkType="signed",power=softPower,minModuleSize=30,
  mergeCutHeight=mergingThresh,numericLabels=TRUE,saveTOMs=TRUE,
  pamRespectsDendro=TRUE,saveTOMFileBase="DilExp",TOMType = 'signed',
  deepSplit = 2)

moduleLabelsAutomatic=net$colors

# Convert labels to colors for plotting
moduleColorsAutomatic = labels2colors(moduleLabelsAutomatic)

# A data frame with module eigengenes can be obtained as follows
MEsAutomatic=net$MEs

# Next use this trait to define a gene significance variable
GS.trait=as.numeric(cor(datExp0,oneTrait,use="p"))
# This translates the numeric values into colors
GS.traitColor=numbers2colors(GS.trait,signed=TRUE)

# Next use this trait to define a gene significance variable
GS.secondTrait=as.numeric(cor(datExp0,secondTrait,use="p"))
# This translates the numeric values into colors
GS.secondTraitColor=numbers2colors(GS.secondTrait,signed=TRUE)

blocknumber=1
datColors=data.frame(moduleColorsAutomatic,GS.traitColor,GS.secondTraitColor)[net$blockGenes[[blocknumber]],]

# Plot the dendrogram and the module colors underneath
plotDendroAndColors(net$dendrograms[[blocknumber]],colors=datColors,
  groupLabels=c("Module colors","GS.trait1","GS.trait2"),dendroLabels=FALSE,
  hang=0.03,addGuide=TRUE,guideHang=0.05)


```


Or do this manually...checking parameters for each stage...

```{r TopologicalOverlapMatrix,eval= TRUE}

#calculate the adjacency matrix [means of representing which vertices (or nodes) of
#a graph are adjace to which other vertices]. See this web page for a nice description of adjacency: http://en.wikipedia.org/wiki/Adjacency_matrix
adjacency = adjacency(datExp0,power = softPower,type = 'signed',corFnc = 'cor')

#topological overlap matrix is essentially a measure of the interconnectedness of each item
#to convert the adjacency matrix into a dissimilarity based on the topological
#overlap...use the TOMdist command
dissTOM = TOMdist(adjacency,TOMType = 'signed')

geneTree = flashClust(as.dist(dissTOM),method = "average")
# sizeGrWindow(12,9)
plot(geneTree,xlab="",sub="",main = "gene cluster on TOM-based dissimilarity",
     labels = FALSE,hang = 0.04)


```


```{r cutTree, eval = TRUE}
#now...do the dynamic branch cutting. Using two branch cutting methods
#to use the hybrid method need both the gene tree and the dissimilarity measure
#the deepSplit parameter dictates how the branches are cut (2,3 gives smaller modules; 
#1 give bigger modules)
##might be better to err on the side of larger modules, especially if there is
##noise in the dataset (acc. to Horvath)
#set the minimum size of the module (30 was suggested in the tutorial)
minModuleSize = 30

# #module identification using dynamic tree cut
##doing the hybrid cutting (and relying on dissTOM) can bad results with groups spanning
##different branches of the tree (and differs from the automatic options)
dynamicMods = cutreeDynamic(dendro = geneTree,
                            method = 'tree',cutHeight = 0.995,
                            deepSplit = 2, pamRespectsDendro = FALSE,
                            minClusterSize = minModuleSize)

table(dynamicMods)

#convert the numeric labels to colors
dynamicColors = labels2colors(dynamicMods)
table(dynamicColors)

#plot the dendrogram, with the colors under it
# sizeGrWindow(8,6)
plotDendroAndColors(geneTree,dynamicColors,"DynamicTreeCut",
                    dendroLabels= FALSE,hang = 0.03,
                    addGuide = TRUE,guideHang =0.05,
                    main = "gene dendrogram and module colors")

# # Relabel the manual modules so that their labels
# # match those from our previous analysis
# moduleLabelsManual2=matchLabels(dynamicMods,moduleLabelsAutomatic)
# # Convert labels to colors for plotting
# moduleColorsManual2=labels2colors(moduleLabelsManual2)

```

Want to merge some modules with similar expression. First, to quantify co-expression similarity of entire modules, we calculate their eigengenes and cluster them on their correlation


```{r calculateEigenGenes2, eval=TRUE}

#calculate eigengenes
MEList = moduleEigengenes(datExp0,colors = dynamicColors)
MEs = MEList$eigengenes

MET=orderMEs(cbind(MEs,oneTrait))
# Plot the relationships among the eigengenes and the trait
plotEigengeneNetworks(MET,"",marDendro=c(0,4,1,2),
    marHeatmap=c(3,4,1,2),cex.lab=0.8,xLabelsAngle=90)

# #calculate dissimilarity of module eigengenes
# MEDiss = 1-cor(MEs)
# 
# #cluster module eigengenes
# METree = flashClust(as.dist(MEDiss),method = "ward")
# 
# #plot the results
# # sizeGrWindow(7,6)
# plot(METree,main = "clustering of module eigengenes",
#      xlab = "",sub = "")
 
# abline(h=mergingThresh,col = "red") #plot the cut line

```

```{r mergeModules, eval=TRUE}

#call an automatic merge function
merge = mergeCloseModules(datExp0,dynamicColors,MEs = MEs,
                                        cutHeight = mergingThresh,verbose=3,
                                        corFnc = 'cor')

mergedColorsManual3 = merge$colors #set new colors
 
# #make better colors...use RColorBrewer 
# #these look nice, but then give me colors as hexadecimal
# cols = colorRampPalette(brewer.pal(12,"Set3"))(length(unique(merge$colors)))
# mergedColorsManual3 = labels2colors(merge$colors,col = cols)

#this are the eigengenes of the new merged modules
MEsManual = merge$newMEs

#plotting
# sizeGrWindow(12,9)
datColors = data.frame(moduleColorsAutomatic,dynamicColors,mergedColorsManual3,GS.traitColor)

plotDendroAndColors(geneTree,colors = datColors, 
            c("Automatic","first Attempt","manual hybrid","GS.trait"),
            dendroLabels = FALSE, hang = 0.03,
            addGuide = TRUE,guideHang = 0.05)
                    
# #rename the colors so we use the new list of colors (from the merged modules)
# moduleColors = mergedColors
# colorOrder = c("grey",standardColors(50)) #color 'grey' is for unassigned genes
# moduleLabels = match(moduleColors,colorOrder)-1
# MEs = mergedMEs

```

Now move on to relate the various modules defined above with the environmental data I have collected for the samples. From the original paper: "since we have a summary profile (eigengen) for each module, we simply correlate eigengenes with external traits and look for the most significant associations"


```{r relateToEnvironmentalParameters, eval=TRUE}

#define the number of genes and samples
nGenes = ncol(datExp0)
nSamples = nrow(datExp0)

#recalculate the MEs with color labels; where ME is 'module eigengene' (1st principal component) of modules in a given single dataset
MEs0 = moduleEigengenes(datExp0,mergedColorsManual3,excludeGrey=FALSE)$eigengenes
#as a test, try the automatic colors
#MEs0 = moduleEigengenes(datExp0,moduleColorsAutomatic,excludeGrey=FALSE)$eigengenes

MEs = orderMEs(MEs0)
moduleTraitCor = cor(MEs,datTraits,use="p")
moduleTraitPvalue = corPvalueStudent(moduleTraitCor,nSamples)

#plot up the results with the appropriate color coding
# sizeGrWindow(9,6)
textMatrix = paste(signif(moduleTraitCor,2),"\n(",
                   signif(moduleTraitPvalue,1),")",sep = "")

dim(textMatrix) = dim(moduleTraitCor)

windows(15,10) #this will make the figure window appear outside the bounds of the RStudio stuff (and hence be easier to edit)
par(mar = c(15,9,2,3))
labeledHeatmap(Matrix = moduleTraitCor,xLabels = names(datTraits),yLabels=names(MEs),
  ySymbols = names(MEs),colorLabels = FALSE, colors = blueWhiteRed(50),
  textMatrix =  textMatrix,xLabelsAngle=45,
  setStdMargins = FALSE, cex.text = 0.5,zlim = c(-1,1),main = paste("module-traits"))

```

```{r getSignificant, eval = TRUE}

#show me the list of compounds that are significantly correlated to a color module

#convert to a dataframe in order to retain the row name information
fC = 'MEmidnightblue'
showC <- data.frame(cbind(moduleTraitCor[fC,],moduleTraitPvalue[fC,]))
showC[showC$X2<0.05,]


```


```{r findInteresting, eval = TRUE}

##one example: 
# calculate the module membership values (aka. module eigengene based connectivity kME):
#find interesting 'genes', now have a data frame of kME values (datKME)
datKME=signedKME(datExp0, MEs)
colorOfColumn=substring(names(datKME),4)
par(mfrow = c(2,2))

selectModules=c("black","blue")
par(mfrow=c(2,length(selectModules)/2))

for (module in selectModules) {
  column = match(module,colorOfColumn)
  restModule=mergedColorsManual3==module
  verboseScatterplot(datKME[restModule,column],GS.trait[restModule],
    xlab=paste("Module Membership ",module,"module"),ylab="GS.weight",
    main=paste("kME.",module,"vs. GS"),col=module)}

```


Next section: gene relationship to trait and important modules: gene significance and module membership.

We quantify associations of individual genes with our trait of interest by defining Gene Significance (GS) as (the absolute value of) the correlation between the geen and the trait. For each module, we also define a quantitative measure of module membership MM as the correlation of the module eigengene and the gene expression profile. This allows us to quantify the similarity of all genes on the array to every module.


```{r nextThing,eval=TRUE}

#set the names (colors) of the modules
modNames = substring(names(MEs),3)
geneModuleMembership = as.data.frame(cor(datExp0,MEs,use = "p"))
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership),nSamples))

names(geneModuleMembership) = paste("MM",modNames,sep="")
names(MMPvalue) = paste("p.MM",modNames,sep = "")

geneTraitSignificance = as.data.frame(cor(datExp0,oneTrait,use = "p"))
GSPvalue = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificance),nSamples))

names(geneTraitSignificance) = paste("GS.",names(oneTrait),sep="")
names(GSPvalue) = paste("p.GS.",names(oneTrait),sep="")

```

Find genes with high significance for weight and high module membership in interesting modules. 

```{r identifyHighGSandMM, eval = TRUE}

#find a hub, gene with high connectively (random subset of genes searched)
chooseOneHubInEachModule(datExp0,mergedColorsManual3)

#find the gene with the highest connectivity (search all genes)
chooseTopHubInEachModule(datExp0,mergedColorsManual3)


```

We now have modules with high association with a specific trait, and identified central players based on module membership. Merge this information with the gene annotation (metabolite info) and write out a file that summarizes the results...


```{r prepareExport, eval = TRUE}
#create the data frame
geneInfo0 = data.frame(nFeature = probes,
  moduleColor = mergedColorsManual3,
  geneTraitSignificance,
  GSPvalue)

#order modules by weight
modOrder = order(-abs(cor(MEs,oneTrait,use = "p")))

#add module membership information in chosen order
for (mod in 1:ncol(geneModuleMembership))
  {
  oldNames = names(geneInfo0)
  geneInfo0 = data.frame(geneInfo0,geneModuleMembership[,modOrder[mod]],
                         MMPvalue[,modOrder[mod]])
  names(geneInfo0) = c(oldNames,paste("MM.", modNames[modOrder[mod]],sep=""),
                       paste("p.MM",modNames[modOrder[mod]],sep=""))
  }

# #order the genes in the geneInfo variable first by module color and then by
# #geneTraitSignificance
# geneOrder = order(geneInfo0$moduleColor,-abs(geneInfo0$GS.caffeine))
# geneInfo = geneInfo0[geneOrder,]

#add the mz and RT information to the last columns:
gI <-match(geneInfo0$nFeature,annot$feature)
geneInfo0$mz <-annot$mz[gI]
geneInfo0$RT <-annot$RT[gI]
geneInfo0$withMS2 <-annot$withMS2[gI]

#this sends out the information for each mzRT feature
write.csv(geneInfo0,file = WriteCSVname)

#this is the file with the information for each color(use to make pretty table)
write.csv(table(mergedColorsManual3),file = writeColorsName)

```

Output some of the modules to Cytoscape for plotting

```{r exportCyto, eval = TRUE}

#select modules
#modules = "darkgreen" # just one module...or two in line below
#modules = c("lightgreen")
modules = modNames #select all modules

#select module probes
probes = names(datExp0)
inModule = is.finite(match(mergedColorsManual3,modules))
modProbes = probes[inModule]

modGenes = annot$mz[match(modProbes,annot$feature)]


#select the corresponding TOM
TOM = TOMsimilarityFromExpr(datExp0,power = softPower,corType = 'pearson',networkType = 'signed',
                            TOMType = 'signed')

modTOM = TOM[inModule,inModule]
dimnames(modTOM) = list(modProbes,modProbes)

#make better colors for Cytoscape:
#pretty, but hard to see
#cols = colorRampPalette(brewer.pal(11,"PRGn"))(length(unique(merge$colors)))


#use threshold to raise/lower the number of interactions we actually want to consider. Was 0.3 as the setting from the tutorial.
cyt = exportNetworkToCytoscape(modTOM,
     edgeFile = paste("Cytoscape-edges_module_negOmni_neg",".txt",sep = ""),
     nodeFile = paste("Cytoscape-nodes_module_negOmni_neg", ".txt",sep = ""),
     weighted=TRUE,
     threshold = 0.3,
     nodeNames = modProbes,
     altNodeNames = modGenes,
     nodeAttr = mergedColorsManual3[inModule])


```


```{r saveStuff, eval = TRUE}

save.image(NameOfFile)

```

