Processing the Dilution Experiment samples. More robust peak shape requirements
FT-LC data, negative ion mode, intracellular metabolites only. 
Use this file and parameters to convert mzML data into a list of features from the 
untargeted metabolomics data.
Krista Longnecker
Woods Hole Oceanographic Institution
July 29, 2015; April 20, 2016; May 20, 2016
========================================================

Go ahead and load up some of the parameters:
```{r Settings, eval=TRUE}
library(xcms)
library(CAMERA)
library(XLConnect)

```


Now go ahead and do the peak picking, or load in the existing file...peak picking can be time consuming and it easier to only do that once

```{r GettingData, eval = TRUE}

##Note: in negative ion mode, I set a threshold for peaks in MSconvert before I even get to this step. This massively improves the peak picking in XCMS
#data are LC data from a 7T FT-ICR-MS instrument. LC parameters given in Kido Soule et al. (2015, Marine Chemistry)
usePath <- "C:/pathToData/FTLCdata/mzML_threshold1000/neg"

ext <- ".mzML"
pre <- paste0(usePath,"/")

mzdatafiles <- list.files(usePath,recursive = FALSE, full.names=TRUE,
                          pattern = paste0("\\",ext))

## How many CPU cores has your machine (or cluster) ?
nSlaves=4

#For negative ion mode: ppm = 2,3 seems best
xs<-xcmsSet(mzdatafiles, method = "centWave",ppm = 2 ,snthresh = 10,
  prefilter = c(3,500), mzCenterFun = "wMean",integrate = 2, 
  verbose.columns = TRUE, peakwidth=c(20,80), fitgauss= TRUE, noise = 500, 
  mzdiff=-0.005,nSlaves=nSlaves)

```


```{r settingFileNames}
# #setup the filenames all in one place bc easier...
#put this here bc if we load in data already with peak picking...that will bring in a NameOfFile

nDate <- "2016.05.20_v1"
nBase <- "DilExp_neg_Omni"
fName_allPeaks <- paste0(nBase,'_listAllPeaks.',nDate,'.csv')
fName_aligned <- paste0(nBase,'_aligned.',nDate,'.csv')
fName_peaksPicked <- paste0(nBase, '_picked.',nDate, '.pdf')
NameOfFile <- paste0(nBase,"_",nDate,".Rimage")
rm(nDate,nBase)

```



```{r apply_peakShape, eval=TRUE}
# # #####the peakShape code seems to be from Tony Larson...via the Google Groups forum  
source("peakShape.r")  

if(1){
  #run peak shape...
  xs2 <- peakShape(xs,cor.val=0.95)
  } else {
    #don't bother
    xs2<-xs
}


```


Setup the ability to rename the classes based on information in the CSV file

```{r getSampleInformation, eval=TRUE}
# use this to match the files with the information about the samples
wDir <-  "pathToFiles/sequenceFiles/"
all <- readWorksheetFromFile(paste0(wDir,"DilutionExperiment_FTLCsequenceInformation.xlsx"),sheet = 1)
putDataHere <-data.frame()

for(i in 1:length(all$File.Name)) {
  h <- all$File.Name[i]
  #paste0 will strip out the spaces; make the full filename here    
  h <-paste0(pre,h,ext)
  
  #this will pick up the right row in mzdatafiles
  m <- match(basename(h),basename(xs2@filepaths))
  if (!is.na(m)) {
    tData <- all[i,]
    if (nrow(putDataHere)==0) {
      putDataHere <- tData
      } else {
    putDataHere[m,] <-tData
    }
    rm(tData)
  }   
  rm(h,m)
}   

rm(all) #house cleaning...

# rm(classlist) 
classlist <- matrix()
#this next bit is a hack, but it works in the context of xcms
for (i in 1:nrow(putDataHere)) {
  classlist[i] <- paste0(putDataHere$ionMode[i], ".", putDataHere$GoodData[i],'.', 
                         putDataHere$SampleType[i])
}
rm(i)


```


```{r DoGrouping_1, eval=TRUE}
     
##now I will need to consider the groups. First 'group' is actually the ionization mode
#for simplicity, start by using split to get the files from one ionization mode  
useXS <- xs2

#define the classes after the peak picking...
sampclass(useXS) <- classlist 

#note this will select files in negative ion mode that are marked with '1' in the GoodData column
xsOne <- split(useXS,classlist)[["negative.1.Omni"]]
putDataHere.1 <- split(putDataHere,classlist)[["negative.1.Omni"]]
nSamples<-length(xsOne@filepaths)

classlist.1 <- matrix()
#this next bit is a hack, but it works in the context of xcms
for (i in 1:nrow(putDataHere.1)) {
  classlist.1[i] <-paste0(putDataHere.1$ionMode[i],".",putDataHere.1$GoodData[i],".",
                        putDataHere.1$SampleType[i])
}

#set the groups to the QC/blank/samples before grouping:
sampclass(xsOne) <-classlist.1

##do this two ways...with pre-grouping
xgN3 <-group.density(xsOne,minfrac=0,minsamp=1,bw=22,mzwid=0.015)     

#set the center sample so it doesn't pick the wacky 1 um sample...
rc.obi <-retcor.obiwarp(xgN3,plottype="deviation",profStep = 0.1,distFunc = "cor", 
                         gapInit = 0.4,gapExtend=0.4,response = 1,center = 5)    

xgN3 <-group.density(rc.obi,minfrac=0,minsamp=1,bw=20,mzwid=0.001)     
xgF <- fillPeaks(xgN3, method = "chrom")                              

```

Go ahead and spit out some variables to the folder

```{r SaveData, eval = FALSE}

#this will spit out the list of all features identified
write.csv(xgF@peaks,file = fName_allPeaks)

```

These are the pieces for CAMERA
      
```{r CAMERA, eval = TRUE}
#
#nSamples <- length(mzdatafiles)
xsa<-xsAnnotate(xgF)

#group the features initially just by retention time
xsaF <-groupFWHM(xsa)

#figure out which features also have a matching 13C feature. Have to enter both 
#the relative error (ppm) and the absolute error (mzabs)
xsaFI <-findIsotopes(xsaF,ppm=1.5,mzabs = 0.0001,minfrac = 1/nSamples)

#now group by the correlations based on (1) intensity, (2) EIC, (3) isotopes...
xsaC <-groupCorr(xsaFI,cor_eic_th=0.75,pval=0.05, graphMethod="hcs",
 calcIso = TRUE, calcCiS = TRUE, calcCaS = FALSE)

#setup the file to also look for adducts, only go with the primary adducts for the moment
file <-system.file("rules/primary_adducts_neg.csv",package = "CAMERA")
rules <-read.csv(file)
an <-findAdducts(xsaC,polarity = "negative",rules=rules,ppm=1.5)

#do some housecleaning
rm(xsa,xsaF,xsaFI,xsaC)
```


And save the final steps

```{r lastSave, eval =TRUE}
#export the data to the CSV file
write.csv(getPeaklist(an),file=fName_aligned) 
save.image(NameOfFile)

```

